---
title: "PS270 Final Project"
author: "Annie Yu"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## (1) Introduction 

Mass incarceration and high recidivism rates have defined the American criminal justice system for decades. Expedited by the War on Drugs and perpetuated by the policies and stigmas surrounding incarcerated individuals post-release, it is pertinent to public health that the underlying mechanisms driving such a concern is better understood. Surprisingly, there is a lack of longitudinal research that specifically isolates the relationship between the presence of substance use and the likelihood of recidivism.

Because co-occurring disorders are common, it is difficult to isolate a control group of inmates who have never struggled with substance use and are honest about it. It is even harder to follow this sample across an extended period of time to determine if they re-offended or if they have introduced drugs into their lifestyles. This report seeks to answer the question: how does recidivism vary among those who experience substance use? I take a different route in examining the relationship between substance use and recidivism by focusing on specific crime categories—violent, property, drug, and public order offenses—then analyzing how substance use trends within these categories relate to the tendency towards recidivism. These are the groupings of crime according to the Bureau of Justice Statistics. 

My hypothesis assumes that substance use is an underlying mechanism that influences what crime an individual is likely to commit and therefore how likely recidivism is for them. I come to this conclusion by taking into the lifelong ramifications that substance use has on an individual and its interconnected character that affects all aspects of an individual: physical health, mental health, socioeconomic tendencies, behavior, and cognition. Therefore, I predict that those who struggle with substance use are more likely to commit crimes that are likely to be re-offended, resulting in higher rearrest, reincarceration, and reconviction rates of individuals with problematic substance use. 

The primary explanatory variable is the presence of substance use in an individual's life. This can be measured during the screening and assessment process of intake for correctional facilities, drug testing, self-reported surveys, and medical records. These records alone will show variation among the sample, because those incarcerated now or in the past have different histories with substance use along with varying levels of severity when it comes to disorders. This approach must be longitudinal, because we must see changes or lack of changes in substance use over the span of their life before and during incarceration, along with the three years following their release. That way, we can see at what points in time substance use was at its peak in relation to their criminal behavior. 

The outcome variable of interest is the rate of recidivism for each individual. This is typically measured by the number of re-arrests, re-convictions, and re-incarcerations that occur within three years of being released from jail/prison, including those released on extended supervision/parole. This data is typically public information. 

One observed pattern in the data that would provide support for my hypothesis is higher recidivism rates among those struggling with one or more substance use disorders. 

An observed pattern that would disprove my hypothesis is higher recidivism rates among those without the presence of substance use or those with relatively low levels of substance use. This would mean that substance use is not a causal indicator of high recidivism and that there is a confounding variable that has a larger effect on the likelihood that an individual will re-offend. For instance, mental illness, employment, environment, housing, education, etc.  

## (2) Data 

To give a general idea of the relationship between substance users and the likelihood of recidivism, figure 1 reflects the disparity between the rearrest rates of those without the presence of substance use compared to those who do. The independent, or explanatory variable, the presence of substance use, has been measured using 36 months of post-release records for New Jersey state inmates. Since this type of study uses data over time, it can be best classified as longitudinal and retrospective in nature. The dependent, or outcome, variable has also been measured by using government clinical records. Clinical records has multiple points of sources, likely a combination of screenings, clinical assessments, and documentation from correctional facilities. Statistical analysis also aided the process of organizing the data. 

Data source: Zgoba, K. M., Reeves, R., Tamburello, A., & Debilio, L. (2020). Criminal Recidivism in Inmates with Mental Illness and Substance Use Disorders

```{r}

library(tidyverse) 
library(infer) 

data_1 <- read_csv("/Users/annieyu/Documents/finalprojectdata.csv")

data_long <- data_1 |>
  select(sud_rearrest, no_sud_rearrest) |>
  pivot_longer(
    cols = sud_rearrest : no_sud_rearrest,
    names_to = "variable",
    values_to = "value")

dataplot <- ggplot(data_long, 
                   aes(x = variable,
                       y = value,
                       fill = variable)) +
  geom_bar(stat = "identity") +
  labs(title = "Figure 1: Rearrests Among Those With/Without Presence of Substance Use", 
       x = "Sample Groups", 
       y = "Rearrest Rate (%)",
       caption = "Data source: Zgoba, K. M., Reeves, R., Tamburello, A., & Debilio, L. (2020). Criminal Recidivism in Inmates with Mental Illness and Substance Use Disorders ",
       fill = "Sample Groups") +
  ylim(0,80) + 
  scale_x_discrete(labels = c(
    "sud_rearrest" = "Substance Use Present",
    "no_sud_rearrest" = "No Substance Use")) +
  scale_fill_manual(values = c(
    "sud_rearrest" = "#1f78b4",
    "no_sud_rearrest" = "#33a02c"),
    labels = c(
      "sud_rearrest" = "Presence of Substance Use On Release",
      "no_sud_rearrest" = "No Presence of Subtance Use On Release")) +
  theme_minimal() +
  theme(panel.background = element_blank(),
        plot.caption = element_text(hjust = 0.1, size = 5),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.1),
    axis.title = element_text(size = 12, face = "bold")
  )


dataplot
```

## (3) Results

The second set of data in Figure 2 comes from the New Survey of Prison Inmates Data Analysis Tool (SPI DAT) from the Bureau of Justice Statistics. It's a dynamic tool that allows public access to interact with data visualizations, with custom charts for a range of characteristics of incarcerated individuals in 2016. The data comes from a national survey of prisoners aged 18 or older in both state and federal correctional facilities. Prisoners with a substance use disorder (SUD) fulfill the criteria for the DSM-IV for either dependence or abuse of alcohol or drugs 12 months prior to incarceration. 

```{r}
sud_by_offense <- read_csv("/Users/annieyu/Documents/ps270finalproject/sud_by_offense_2016.csv")

off_data <- sud_by_offense |>
  mutate(SUD = 
           if_else( sud == "Yes", 1, 0))
knitr::kable(head(off_data))

off_data_plot <- off_data |>
  ggplot(aes(x = offense_type,
             y = Perc, 
             fill = as.factor(SUD))) + 
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("0" = "#33a02c",
                               "1" = "#1f78b4"),
                    labels = c("No SUD", "Yes SUD")) +
  scale_x_discrete(labels = c(
                               "drug" = "Drug",
                               "other" = "Other",
                               "property" = "Property",
                               "public_order" = "Public Order",
                               "violent" = "Violent")) +
  ylim(0,100) +
  labs(
    title = "Figure 2: Types of Offenses and SUD Presence in Offenders",
    x = "Type of Offense",
    y = "Percentage of Offenses Committed",
    fill = "Offender SUD Status",
    caption = "Data source: New Survey of Prison Inmates Data Analysis Tool (SPI DAT) from the Bureau of Justice Statistics") +
  theme_minimal() +
  theme(panel.background = element_blank(),
        plot.caption = element_text(hjust = 0.5, size = 5),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.1),
    axis.title = element_text(size = 11, face = "bold"))

off_data_plot
```

This visualization suggests that property crimes are more likely to be committed by those with a substance use disorder than those without. 

Regression 

This third data data set comes from the Wisconsin Department of Corrections. The recidivism calculations of rearrests contain 3 components: report of release date, a 3-year follow-up period, and the documentation of a recidivism event. The sample is all persons released from prison in Wisconsin from 2000-2016.

```{r}
raoodata <- read_csv("/Users/annieyu/Documents/ps270finalproject/rearrest_by_original_offense.csv")

head(raoodata)

raoodata_summary <- raoodata |>
  summarize(
    "Drug Offenses" = mean(total_recid_drug_perc * 100),
    "Property Offenses" = mean(total_recid_prop_perc * 100),
    "Public Order Offenses" = mean(total_recid_pub_perc * 100),
    "Violent Crime Offenses" = mean(total_recid_vio_perc * 100)) |>
  pivot_longer(
    cols = everything(), 
    names_to = "offense_type", 
    values_to = "rearrest_rate") |>
  rename("Offense Type" = offense_type,
         "Rearrest Rate" = rearrest_rate)

raoodata_sum_table <- raoodata_summary |>
  knitr::kable(digits = 2, caption = "Average Rearrest Rates by Offense Type")
raoodata_sum_table
```

```{r}
rearrest_long <- raoodata |>
  pivot_longer(
    cols = starts_with("total_recid"),
    names_to = "offense_type",
    values_to = "rearrest_rate") |>
  mutate(crime_type = case_when(
    offense_type == "total_recid_drug_perc" ~"Drug",
    offense_type == "total_recid_prop_perc" ~ "Property",
    offense_type == "total_recid_pub_perc" ~ "Public Order",
    offense_type == "total_recid_vio_perc" ~ "Violent"),
    control_group = if_else(
      offense_type == "Violent", 1, 0))
knitr::kable(head(rearrest_long, 5))
```

Bootstrapping Model

As shown in the visualization, the bootstrapping model isn't very compatible with the data set, because the sample size is not extensive enough to generate a solid null distribution. An F-statistic is being used instead of difference in proportions or difference in means, because I am trying to measure the variability between the rearrest rates of 4 types of crime. Additionally, I am trying to evaluate whether the difference in the average rearrest rates are statistically significant compared to the overall variability in the data. The dashed line shows the standard deviation of the distribution and the black line represents the density curve of the permutations.  

```{r}
null_dist <- rearrest_long |>
  specify(rearrest_rate ~ crime_type) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "F")

observed_stat <- rearrest_long |>
  specify(rearrest_rate ~ crime_type) |>
  calculate(stat = "F")

ci_95 <- null_dist |>
  get_confidence_interval(level = 0.95, type = "percentile")

rearrest_pval <- null_dist |>
  get_p_value(obs_stat = observed_stat, direction = "greater")

knitr::kable(head(null_dist, 5))
knitr::kable(observed_stat)
knitr::kable(ci_95)
knitr::kable(rearrest_pval)
```

```{r}
null_dist_plot <- null_dist |>
  ggplot(aes(x = stat)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 0.2, fill = "grey") +
  geom_density(color = "#1f78b4", size = 1) +
  stat_function(
    fun = dnorm,
    args = list(mean = mean(null_dist$stat), sd = sd(null_dist$stat)),
    color = "black",
    linetype = "dashed",
    size = 1
  ) +
  labs(
    title = "Bootstrapping Null Distribution",
    x = "F-Statistic",
    y = "Density"
  ) +
  theme_minimal() +
  theme(panel.background = element_blank(),
        plot.caption = element_text(hjust = 0.1, size = 5),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.1),
    axis.title = element_text(size = 12, face = "bold")
  )

null_dist_plot
```

Linear Regression of Rearrests

```{r}
rearrest_long <- rearrest_long |>
  mutate(crime_type = factor(crime_type, levels = c("Violent", "Drug", "Property", "Public Order")))

lm_rearrest <- lm(rearrest_rate ~ crime_type, data = rearrest_long)

modelsummary::modelsummary(lm_rearrest, statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared","adj.r.squared"))
```

The regression table shows that when $\alpha$ = 0.05, property offenses significantly predict higher rearrest rates, with an increase of 12.31 percentage points compared to violent offenses. Violent offenses was used as a reference group, since it had the lowest rearrest rate of the 4 categories. Drug offenses and public order offenses also predict higher rearrest rates, but only with increases of 5.45 and 3.69 percentage points. This suggests that property offenses have the strongest association with rearrest rates among these four main crime categories. The p-value of <0.001 means that we can reject the null hypothesis that there is no significant association between one's original offense of a property crime and the likelihood of rearrest within 3 years post-release. 

This analysis only shows a correlation, not a causal effect. Demographic variables have not been accounted for because that information is not included in publicly available data. 

## (4) Conclusion

This analysis provides some insight into what types of crime those with substance use disorders commit and if the type of crime committed has an effect on their likelihood of rearrest. I use violent crime as a reference group, since those with substance use disorders make up a smaller percentage of offenders compared to those without. This may have implications on the analysis, because it is not a perfect control group.

I recognize that this analysis is very far from providing a causal explanation on the relationship between substance use and recidivism, but I believe it provides a unique perspective on the character of those within the criminal justice system who struggle with substance use. Although the regression points to the conclusion that those whose original offense was a property crime may have a higher likelihood of rearrest than those with different offenses, it doesn't include substance use in the calculation. Rather, it is assumed the majority of people in the sample who commit property crimes are people with a SUD. Therefore, the hypothesis isn't directly answered due to such limitations in the data. The largest obstacle in conducting this report is the lack of access to one comprehensive longitudinal data set that follows an individual after release from incarceration. There are only bits and pieces of data available to the public for confidential reasons. Confounding variables such as age, gender, race, prior offenses, sexuality, socioeconomic status, etc. are not accounted for. Following the intersectional framework of sociology, these overlapping identities are very important to how one experiences daily life and what disadvantages they may be prone to. My report merely highlights the correlation that property crimes are more likely to be committed by those with a substance use disorder and those whose original offense is a property crime are more likely to be rearrested 3 years post-release.

Another important consideration is that this analysis excludes the other two categories of recidivism: reincarceration and reconviction. I chose to use rearrests as the primary baseline of data, because a rearrest is simply the arrest of an individual who has committed a criminal act. Reincarceration and reconviction data provide insight into what happens to the individual after charges are made. These two factors are not incredibly important to the hypothesis, because whether or not someone is reincarcerated and reconvicted depends on a multitude of variables that are difficult to measure: the discretion of the prosecutor, law enforcement, judge, and probation/parole officer (if applicable), the rehabilitation programs available in the district, biases, etc. The important matter is whether or not the individual has behaved in a way that warrants a documented arrest. 


## Works Cited

https://spi-data.bjs.ojp.gov/

https://doc.wi.gov/DataResearch/RecidivismReincarceration/0821RecidivismAfterReleaseFromPrison.pdf




